Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-07-28 paper/main.tex
related work equiv

\section{Related Work}
We build on the extensive work codifying the structure of visualization data to develop a more generalizable method for expressing the structure that visualizations are expected to preserve. Broadly, previous work describes structure as a combination of the topological properties of the data, the mathematical structure of the data fields, and equivalent changes to data and graphics. We propose that a unified abstraction model can guide developers in building structure aware composable, reusable software library components with more consistent interfaces.

@@

\subsection{Data Field Structure}
\label{sec:related-work:equivariance}
Along with preserving topology, visualization algorithms are expected to preserve the structure of the data fields, i.e. the columns in a table, when they translate data to visual representations. The field structure of data has traditionally been described using the measurment scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. The expectation that visual transformation is one where the measurement structure be matched by the visual encoding was codified by Bertin \cite{bertinSemiologyGraphicsDiagrams2011}. An \textit{equivariant} map is the formal term for a function where the input and output have symmetric group structure. Given a group $G$ that acts on input $X$ and output $Y$ and a function $f: X \rightarrow Y$, the function $f$ is \textit{equivariant} if $f(g(x)) = g(f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant}. For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. Mackinlay\cite{mackinlayAutomaticDesignGraphical1987} more generally definines visualization transformations as \textit{homomorphic} maps. A map $f: X \rightarrow Y$ is a \textit{homomorphism} when $X$ and $Y$ have the same algebraic structure type and $f(x_1 \oslash x_2) = f(x_1) \oslash f(x_2)$ for every pair $x_1, x_2$ in $X$\cite{grimaldiDiscreteCombinatorialMathematics2006}. Mackinlay's construction encompasses non-invertable structures such as partial orderings. In \autoref{sec:artist:equivariant:data}, we will also describe the structure of the data in terms of the functions acting on the data.

Structure preservation is important because it means visualzations are easier to understand \cite{norman_things_smart} and ensures that the visual representations are not distortions of the data\cite{tufteVisualDisplayQuantitative2001}. Kindlemannn and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014} propose three specific types of structure preservation: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, that changing the data should correspond to changes in the visualization in a perceptually significant manner. In \autoref{sec:atct:sheaves} we introduce a uniform abstract data representation layer, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the definition of equivariance introduced in \autoref{sec:artist:equivariant:artist} includes the correspondence between data and visual changes. In this work, we formally specify the conditions necessary to build structure preserving components using category theory, because it provides a rich language for expressing the structure of objects, functions between objects, and the constraints that must hold for these functions to compose \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

>>>>

\section{Related Work}
We build on the extensive work codifying the structure of visualization data to develop a more generalizable method for expressing the structure that visualizations are expected to preserve. Structure preservation is important because it means visualzations are easier to understand \cite{norman_things_smart} and ensures that the visual representations are not distortions of the data\cite{tufteVisualDisplayQuantitative2001}. Broadly, previous work describes structure as a combination of the topological properties of the data, the mathematical structure of the data fields, and equivalent or compatible changes to data and graphics. We propose that a unified abstraction model can guide developers in building structure aware composable software library components with more consistent interfaces.

@@

\subsection{Data Field Structure}
\label{sec:related-work:equivariance}
Along with preserving topology, visualization algorithms are expected to preserve the structure of the data fields, i.e. the columns in a table, when they translate data to visual representations. 
The field structure of data has traditionally been described using the measurment scales-i.e. 
nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. The expectation that visual transformation is one where the measurement structure be matched by the visual encoding was codified by Bertin \cite{bertinSemiologyGraphicsDiagrams2011}. Two mathematical concepts for expressing structure preservation are \textit{equivariance} and \textit{homomorphism}:
\begin{itemize}
\item an \textit{equivariant} function preserves symmetric group structure on its input and output. Given a group $G$ that acts on input $X$ and output $Y$ and a function $f: X \rightarrow Y$, the function $f$ is \textit{equivariant} if $f(g(x)) = g(f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant}.
\item a \textit{homomorphic} function preserves binary operations on input and output of the same algberaic type. A map $f: X \rightarrow Y$ is a \textit{homomorphism} when $X$ and $Y$ have the same algebraic structure type and $f(x_1 \oslash x_2) = f(x_1) \oslash f(x_2)$ for every pair $x_1, x_2$ in $X$\cite{grimaldiDiscreteCombinatorialMathematics2006}.
\end{itemize}
For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. Partial orders are non-invertable, so a visual encoding is homomorphic when data that is partially ordered is mapped to visual encodings that are partially ordered in the same way.Mackinlay\cite{mackinlayAutomaticDesignGraphical1987} proposes that data to visual encoders should be \textit{homomorphic} maps in his specification of \textit{A Presentation Tool}. The notion that visual transforms are equivariant underlies Kindlemannn and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}'s Algebraic Visualization Design. They propose three specific types of structure preservation: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, and that changing the data should correspond to changes in the visualization in a perceptually significant manner. In \autoref{sec:atct:sheaves} we introduce a uniform abstract data representation layer, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the definition of equivariance introduced in \autoref{sec:artist:equivariant:artist} includes the correspondence between data and visual changes. In this work, we formally specify the conditions necessary to build structure preserving components using category theory, because it provides a rich language for expressing the structure of objects, functions between objects, and the constraints that must hold for these functions to compose \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.
